Project Diary

### 2014-05-06 thesis/cross_correlation_gaussian.tex
Downsized to 49 pages in the main text. Many images are moved to the appendix.

\section{Distribution of the Eigenvalues}\label{sec:GCC-numerical}
For the Wishart matrix, theoretical results are available for the
eigenvalue distribution \cite{Chiani2012}. In the simplest case where
$\Sigma = I$, i.e. no cross-correlations are present, we have
\begin{eqnarray*}
  f_{\lambda}(x_1, \cdots, x_N) = K \prod_{i=1}^N e^{-x_i/2}
  x_i^{(T-N-1)/2} \prod_{i<j}^N (x_i - x_j)
\end{eqnarray*}
where the eigenvalues have been indexed in descending order: $x_1 \geq
x_2 \geq \cdots \geq x_N$. The normalization constant K is given by
\begin{equation*}
  K = {\pi^{N^2/2} \over 2^{NT/2}\Gamma_N(T/2) \Gamma_N(N/2)}
\end{equation*}
where the function $\Gamma_m(a)$ is defined as
\begin{equation*}
  \Gamma_m(a) = \pi^{m(m-1)/4} \prod_{k=1}^m \Gamma\left(a - {k-1
      \over 2}\right)
\end{equation*}
However, as detailed in the derivation leading to equation
\ref{eq:cross-corr-matrix-PDF}, the distribution of $RR'$ is not
Wishart when the columns of $R$ are correlated. Deriving the
eigenvalue distribution analytically in this case is difficult and
beyond the scope of this thesis. Instead we resort to numerical
methods.

As before we consider the AR(1) process (see \S
\ref{sec:FundamentalConcepts}):
\begin{equation*}
  \vec{r}_t = \phi \vec{r}_{t-1} + \vec{a}_{t}
\end{equation*}
where $\vec{a}_t \sim N(0, I)$, i.e. the elements of $\vec{a}_t$ are
independent Gaussian random variable with zero mean and unit
variance.

Now we investigate how the eigenvalue distribution depends on
$\phi$. Figure \ref{fig:GaussianMarkovSpectrumPDF} shows the results
of the simulation.
% \begin{wrapfigure}{r}{0.5\textwidth}
\begin{figure}[htb!]
  \begin{center}
    \includegraphics[scale=0.5, clip=true, trim=90 228 115
    226]{../pics/GaussianMarkovSpectrumPDF.pdf}
  \end{center}
  % \vspace{-10mm}
  \caption{\small \it
      Eigenvalue distribution with $\tau$ ranging from 0 to
      3. The 1st blue line, which is shown as stairs, is the
      theoretical eigenvalue distribution according to the
      Marcenko-Pastur law: $f(\lambda) = \sqrt{(\lambda_+/\lambda - 1)(1
        - \lambda/\lambda_-)}/2\pi q \sigma^2$, where $\lambda_{\pm} =
      \sigma^2(1 \pm \sqrt{q})^2$ In the simulation we have
      chosen $q = N/T = 50/1000 = 0.05$ and $\sigma=1$. For each value
      of $\tau$ we generate 2000 instances of $N \times T$ random matrix
      $R$, and compute C as $C=RR'/T$. Hence each curve in the figure is
      constructed from 2000 sets of eigenvalues.
    }
  \label{fig:GaussianMarkovSpectrumPDF}
\end{figure}
It is clear from the figure that the maximum eigenvalue moves
consistently to the right as the value of $\phi$ increases. The change
of the minimum eigenvalue remains unclear from the figure, so we plot
in \ref{fig:Gaussian_mineig} their mean for each value of
$\phi$. There one can see that the minimum eigenvalue actually
increases with $\phi$.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5, clip=true, trim=100 226 116
  133]{../pics/Gaussian_mineig.pdf}
  \caption{\small \it The the minimum eigenvalue versus
    auto-correlation strength $\phi$. For each value of $\phi$ 2000
    random matrices are generated and their eigenvalues are
    calculated. The minimum eigenvalue of each random matrix is noted
    and the mean of the 2000 such minimum eigenvalues are plotted
    against the chosen value of $\phi$. 20 values of $\phi$ are
    included in the plot, ranging from 0 to 0.95 with step size 0.05.}
  \label{fig:Gaussian_mineig}
\end{figure}

K. Johnsson \cite{Johnsson2000} and I. Johnstone \cite{Johnstone2001}
showed that, at the {\it absence} of autocorrelations and in the
asymptotic limit $N, T \to \infty$, $N/T \to q < \infty$, the maximum
eigenvalue  $\lambda_1$ follows the Tracy-Widom distribution (denote
$\mathscr{TW}_1$ here) when properly relocated and rescaled:
\begin{equation*}
  {\lambda_1 - \mu_{NT} \over \sigma_{NT}} \sim \mathscr{TW}_1
\end{equation*}
where $\mu_{NT}$ and $\sigma_{NT}$ are given by
\begin{eqnarray*}
  \mu_{NT} &=& \left(
    \sqrt{N-1/2} + \sqrt{T - 1/2}
  \right)^2 \\
  \sigma_{NT} &=& \sqrt{\mu_{NT}} \left(
    {1 \over \sqrt{N-1/2}} + {1 \over \sqrt{T-1/2}}
  \right)^{1/3}
\end{eqnarray*}
So in the following we focus on the maximum eigenvalue distribution
and leave the minimum eigenvalue for future studies. The
$\mathscr{TW}_1$ distribution has the following cummulative
distribution function (CDF) \cite{Chiani2012}
\begin{equation*}
  F_1(x) = \exp\left[
    -{1 \over 2} \int_x^\infty dy \left(q(y) + (y-x)q^2(y)\right)
  \right]
\end{equation*}
where $q(y)$ is defined as the solution to the Painlev\'e II differential
equation
\begin{equation*}
  q''(y) = yq(y) + 2q^3(y)
\end{equation*}
which is unique when imposing the condition
\begin{equation*}
  q(y) \sim \text{Ai}(y) \text{ as } y \to \infty
\end{equation*}

Then Marco Chiani showed recently that the $\mathscr{TW}_1$ distribution
can be well approximated by a gamma distribution based on his proof
that the exact distribution of the maximum eigenvalue is a mixture of
gamma distributions. Specifically,
\begin{equation}\label{eq:TracyWidom-Gamma}
  {\lambda_1 - \mu_{NT} \over \sigma_{NT}} + \alpha \sim
  \mathscr{G}(k ,\theta)
\end{equation}
where $\mathscr{G}(k, \theta)$ denotes the Gamma distribution with
parameters $k$ and $\theta$ \cite{Chiani2012}. The probability
density function (PDF) of the gamma distribution is given by
\begin{equation*}
  f_\gamma(x; k, \theta) = {1 \over \Gamma(k) \theta^k} x^{k-1}
  e^{-x/\theta}
\end{equation*}
Moreover, the first 3 moments of the distribution are simple:
\begin{eqnarray*}
  \text{mean} &=& k\theta \\
  \text{variance} &=& k\theta^2 \\
  \text{skewness} &=& 2/\sqrt{k}
\end{eqnarray*}

In the following we first verify these results and extend them to the
case {\it with} autocorrelations and then study how the
distribution changes as $\phi$ increases. Figure
\ref{fig:GaussianMarkov005MaxEigCDF_loglog} compares the empirical
maximum eigenvalue CDF with the CDF of a gamma distribution. Each pair
of CDFs correspond to a fixed autocorrelation strength ($\phi$). The
parameters $k$ and $\theta$ of the gamma distribution are fit to
data by matching the 2nd and the 3rd moments of the gamma distribution
to the corresponding moments of the empirical distribution. Then the
parameter $\alpha$ in equation \ref{eq:TracyWidom-Gamma} is chosen to
be
\begin{equation*}
  \alpha = k\theta - \E\left({\lambda_1 - \mu_{NT} \over \sigma_{NT}}\right)
\end{equation*}

\begin{figure}
  \begin{center}
    \includegraphics[scale=0.5, clip=true, trim=0 238 0
    197]{../pics/GaussianMarkov005MaxEigCDF_loglog.pdf}
  \end{center}
  \caption{\small \it Cummulative distribution function (CDF) of the maximum
    eigenvalue ($\lambda_1$). Blue: empirical CDF. Red: CDF of the
    fitting gamma distribution. The curves are plotted on log-log
    scale.}
  \label{fig:GaussianMarkov005MaxEigCDF_loglog}
  %\vspace{-10mm}
\end{figure}

We can see in figure \ref{fig:GaussianMarkov005MaxEigCDF_loglog} that
the gamma distribution fits fairly well to its corresponding maximum
eigenvalue distribution. So we conclude that a gamma distribution not
only approximates the maximum eigenvalue distribution at the absence of
autocorrelations but does so even at the {\it presence} of
autocorrelations. Since the maximum eigenvalue distribution is
characterized by the parmaters $k$, $\theta$, and  $\alpha$, the
influence of the autocorrelations can be characterized by the
dependence of $k$, $\theta$, and  $\alpha$ on $\phi$. In the following
we study these dependences in more details.

Figure \ref{fig:param1_phi_dep} shows how the mean, variance,
and skewness of the fitted gamma distribution, as well as the
relocation parameter $\alpha$ denpend on $\phi$.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=37 240 50
    202]{../pics/param1_phi_dep.pdf}
  \caption{\small \it  the  mean, variance, and skewness of the fitted
    gamma distribution as well as the relocation parameter $\alpha$
    are plotted against the autocorrelation strength $\phi$. 20 values
    of $\phi$ are included in the plot, ranging from 0 to 0.95 with
    step size 0.05.}
  \label{fig:param1_phi_dep}
\end{figure}
Despite the monotonic behavior shown by 3 of these curves, the
relation between each of the aforementioned quantities and $\phi$ is
more complicated than it apears. However, the mean, i.e. $k \theta$ is
an exception: Good support in the data can be found for the following
approximate relation:
\begin{eqnarray}
  k\theta &=& a \tan^2{\pi \phi \over 2} + b\tan{\pi \phi \over 2} +
  c \label{eq:k_theta-phi}
\end{eqnarray}
To verify this relation, we first fit a 2nd order polynomial and
obtain the coefficients $a$, $b$, $c$; then for each data point
$k_n\theta_n$ we solve the quadratic equation
\begin{eqnarray}
  a \tan^2{\pi \phi'_n \over 2} + b\tan{\pi \phi'_n \over 2} + c -
  k_n\theta_n &=& 0\label{eq:k_theta-phi_2}
\end{eqnarray}
for $\tan{\pi \phi'_n \over 2}$. If relation \ref{eq:k_theta-phi} is a
good approximation, a close match between $\tan{\pi \phi'_n \over 2}$
and $\tan{\pi \phi_n \over 2}$ is expected. Figure
\ref{fig:phi_and_roots} plots $\tan{\pi \phi'_n \over 2}$ against
$\tan{\pi \phi_n \over 2}$ and fits a straight line to the data
points. It is seen that the data points lie fairly close to the fitted
line and the line has a slope very close to 1 and an intercept close to
0. This strongly suggests $\tan{\pi \phi'_n \over 2} = \tan{\pi \phi_n
  \over 2}$ and supports the relation \ref{eq:k_theta-phi}.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=37 217 39
    170]{../pics/phi_and_roots.pdf}
  \caption{\small \it Upper plot: $\tan{\pi \phi'_n \over 2}$ against
    $\tan{\pi \phi_n \over 2}$. The fitted line has equation $y_n =
    0.995 \tan{\pi \phi_n \over 2} + 0.0146$. Lower plot: Residuals of
    the linear fit, i.e. $\tan{\pi \phi'_n \over 2} - y_n$. 20 values
    of $\phi$ are included in the plot, ranging from 0 to 0.95 with
    step size 0.05.}
  \label{fig:phi_and_roots}
\end{figure}

>>>>

\section{Distribution of the Eigenvalues}\label{sec:GCC-numerical}
For the Wishart matrix $\mtx{RR'}$, theoretical results are available
for the eigenvalue distribution, and we summarize them in appendix
\ref{sec:wishart_eigen_dist}. In short, the joint probability density
function of the eigenvalues is given by equation
\ref{eq:wishart_eigen_pdf} when neither auto-correlation nor
cross-correlation is present in $\mtx R$. Moreover, the mariginal
distribution of the largest eigenvalue is approximated by a gamma
distribution \cite{Chiani2012}.

However, as detailed in the derivation leading to equation
\ref{eq:cross-corr-matrix-PDF}, the distribution of $RR'$ is not
Wishart when the columns of $R$ are correlated
(auto-correlation). Deriving the eigenvalue distribution analytically
in this case is beyond the scope of this thesis. Instead, we resort to
numerical methods.

As before we consider the AR(1) process (see
\S\ref{sec:FundamentalConcepts}):
\begin{equation*}
  \vec{r}_t = \phi \vec{r}_{t-1} + \vec{a}_{t}
\end{equation*}
where $\vec{a}_t \sim N(0, I)$, i.e. the elements of $\vec{a}_t$ are
independent Gaussian random variable with zero mean and unit
variance.

Now we investigate how the eigenvalue distribution depends on
$\phi$. Figure \ref{fig:GaussianMarkovSpectrumPDF} shows the results
of the simulation. 
It is clear from the figure that the maximum eigenvalue moves
consistently to the right as the value of $\phi$ increases, and, as
shown in figure \ref{fig:Gaussian_mineig}, the minimum eigenvalue
also increases with $\phi$.

% In the following we focus on the maximum eigenvalue, which is shown by
% Chiani to have approximately a gamma distribution \cite{Chiani2012}.
% So we first verify Chiani's result, and then extend it to the
% case {\it with} autocorrelations and study how the
% distribution changes as $\phi$ increases.
Chiani showed that the maximum eigenvalue ($\lambda_1$) had
approximately a gamma distribution \cite{Chiani2012}. So we compare in
figure \ref{fig:GaussianMarkov005MaxEigCDF_loglog} the empirical
cumulative distribution function (CDF) of the maximum eigenvalue 
with the CDF of a gamma distribution. The cases where
auto-correlations are present ($\phi > 0$) have also been included.

It is seen in the figure that gamma distributions with different
parameters fit fairly well in all cases. So we conclude that a
gamma distribution not only approximates the maximum eigenvalue
distribution at the absence of autocorrelations but does so even at
the {\it presence} of autocorrelations.

Since the maximum eigenvalue distribution is characterized by the
parameters $k$, $\theta$, and $\alpha$, the influence of the
autocorrelations can be characterized by the dependence of $k$,
$\theta$, and  $\alpha$ on $\phi$. While these dependences are rather
intricate,
% Figure \ref{fig:param1_phi_dep} shows how the mean, variance,
% and skewness of the fitted gamma distribution, as well as the
% relocation parameter $\alpha$ denpend on $\phi$.
% \begin{figure}[htb!]
%   \centering
%     \includegraphics[scale=0.5, clip=true, trim=37 240 50
%     202]{../pics/param1_phi_dep.pdf}
%   \caption{\small \it  the  mean, variance, and skewness of the fitted
%     gamma distribution as well as the relocation parameter $\alpha$
%     are plotted against the autocorrelation strength $\phi$. 20 values
%     of $\phi$ are included in the plot, ranging from 0 to 0.95 with
%     step size 0.05.}
%   \label{fig:param1_phi_dep}
% \end{figure}
% Despite the monotonic behavior shown by 3 of these curves, the
% relation between each of the aforementioned quantities and $\phi$ is
% more complicated than it apears. However, the mean, i.e. $k \theta$ is
% an exception: 
good support can be found in the data for the following
approximate relation:
\begin{eqnarray}
  k\theta &=& a \tan^2{\pi \phi \over 2} + b\tan{\pi \phi \over 2} +
  c \label{eq:k_theta-phi}
\end{eqnarray}
To verify this relation, we first fit a 2nd order polynomial and
obtain the coefficients $a$, $b$, $c$; then for each data point
$k_n\theta_n$ we solve the quadratic equation
\begin{eqnarray}
  a \tan^2{\pi \phi'_n \over 2} + b\tan{\pi \phi'_n \over 2} + c -
  k_n\theta_n &=& 0\label{eq:k_theta-phi_2}
\end{eqnarray}
for $\tan{\pi \phi'_n \over 2}$. If relation \ref{eq:k_theta-phi} is a
good approximation, a close match between $\tan{\pi \phi'_n \over 2}$
and $\tan{\pi \phi_n \over 2}$ is expected. Figure
\ref{fig:phi_and_roots} plots $\tan{\pi \phi'_n \over 2}$ against
$\tan{\pi \phi_n \over 2}$ and fits a straight line to the data
points. It is seen that the data points lie fairly close to the fitted
line and the line has a slope very close to 1 and an intercept close to
0. This strongly suggests $\tan{\pi \phi'_n \over 2} = \tan{\pi \phi_n
  \over 2}$ and supports the relation \ref{eq:k_theta-phi}.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=37 217 39
    170]{../pics/phi_and_roots.pdf}
  \caption{\small \it Upper plot: $\tan{\pi \phi'_n \over 2}$ against
    $\tan{\pi \phi_n \over 2}$. The fitted line has equation $y_n =
    0.995 \tan{\pi \phi_n \over 2} + 0.0146$. Lower plot: Residuals of
    the linear fit, i.e. $\tan{\pi \phi'_n \over 2} - y_n$. 20 values
    of $\phi$ are included in the plot, ranging from 0 to 0.95 with
    step size 0.05.}
  \label{fig:phi_and_roots}
\end{figure}

### 2014-05-06 thesis/price_models2.tex
Downsized to 49 pages in the main text. Many images are moved to the appendix.

\section{GARCH models} \label{sec:Garch_model}
``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A GARCH(p, q) model is defined by the following
equation system:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu + \epsilon_t \\
    \sigma_t^2 &= \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
    \sum_{i=1}^p \beta_i \sigma_{t-i}^2
  \end{aligned}
\end{equation}
where $\epsilon_t$ is a random variable of a prescribed distribution,
the most common ones being Gaussian, Student's t, Asymmetric Student's
t, etc; $\sigma_t^2$ is the variance of $\epsilon_t$
\cite{Bollerslev86}. The autocorrelation function of $\epsilon_t^2$ is
given by (c.f. \cite{Bollerslev87})
\begin{eqnarray*}
\rho_n &=& \sum_{i=1} ^{p \vee q} (\alpha_i + \beta_i) \rho_{n-i}
\text{ for $n > p$}
\end{eqnarray*}
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. $p \vee q$ denotes the maximum of p and q. From these
equations, it is clear that the partial autocorrelation function cuts
off at $\max(p, q)$.

@@

Thus in practice one can try a few frequencies and compare the
distribution of $r_{t+h, h}/\hat{\sigma}_{t+h, h}$ with the standard
Gaussian. If the two match, $\hat{\sigma}_{t+h, h}$ is a good estimate
of $\sigma_{t+h,   h}$, the volatility of $r_{t+h, h}$.

@@

As another measure of the quality of forecast, the fraction of
``good'' forecasts that differ no more than a given percentage from
the measured value is worth a few words. Table
\ref{tab:nordea_good_percentage} compares this fraction of good
forecasts from the 3 alternatives with respect to different
requirements of goodness.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ &
    ARIMA & GARCH & Sample Mean \\
    \hline
    1\% & 0.1920 & 0.2010 & 0.1843 \\
    \hline
    5\% & 0.7887 & 0.7436 & 0.7229 \\
    \hline
    10\% & 0.9665 & 0.9601 & 0.9369 \\
    \hline
  \end{tabular}
  \caption{\small \it Fraction of ``good'' forecasts as defined by
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ being less than 1\%, 5\% and 10\%.}
  \label{tab:nordea_good_percentage}
\end{table}
Clearly, although ARIMA and GARCH rank differently with respect to
different requirements of goodness, they both out-perform the sample
mean forecast. So the efforts of building models and estimating them
have paid off.

\subsection{Volvo B Log-Volatility}\label{sec:volvo}
In this section we model the log-volatility of {\it Volvo B} 30-minute
returns during the period 2013/10/10 - 2014/04/04. This series
contains 1884 log-volatilities computed using 2-minute returns in each
30-minute interval. Among them we use the first 1507 for model
estimation and the last 377 for forecast and model verification.

@@

Following the aforementioned procedure, i.e. computing the initial
estimate by matching the moments and then refining the estimate by
MLE, we obtain the parameter values for a number of return series,
including the Volvo B 30-minute returns described in
\S\ref{sec:SLV_Symmetric}. In the following, we list the parameter
values and show the comparison between the empirical distribution
functions and their counterparts from the model.

Clearly these figures show a fairly good match of the empirical
distribution functions to their model counterparts. So we are once
again assured of the validity of ARIMA log-volatility models. However,
we also see that the skewness of the returns of Volvo B and Ericsson B
have rather different values when computed from the sample and from
the model. This could be the consequence of the limited sample size or
deficiencies in the estimation procedure described above. We leave
these issues to future studies.

@@

\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    & $\psi$ & $\sigma$ & $\bar{v}$ & $\mu$ \\
    \hline
    Volvo B & -1.5747e-02 & 4.6929e-01 & -6.2304e+00 &
    -5.3040e-05 \\
    Nordea Bank & 1.7234e-02 & 4.2780e-01 & -6.3758e+00 & 7.5230e-05
    \\
    Ericsson B & 1.8486e-02 & 4.4905e-01 & -6.2834e+00 & -9.2817e-05 \\
    \hline
  \end{tabular}
  \caption{\small \it Parameter Values of Selected Assets'
    returns. The time series are 30-minute returns and run from
    2013/10/10 to 2014/03/12.}
  \label{tab:assets_params}
\end{table}

\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    & mean & std & skewness & kurtosis \\
    \hline
    \multirow{2}{*}{Volvo B} & -8.6766e-05 & 2.4500e-03 &
    -2.7217e-02 & 7.3394e+00 \\
    & -6.9281e-05 & 2.4538e-03 & -7.2677e-02 & 7.2472e+00 \\
    \hline
    \multirow{2}{*}{Nordea Bank} & 7.4821e-05 & 2.0500e-03 &
    6.9433e-02 & 6.7427e+00 \\
    & 8.8983e-05 & 2.0443e-03 & 6.7125e-02 & 6.2446e+00 \\
    \hline
    \multirow{2}{*}{Ericsson B} & -8.2606e-05 & 2.2939e-03 &
    2.6586e-01 & 7.6716e+00 \\
    & -7.5675e-05 & 2.2844e-03 & 7.8573e-02 & 6.7299e+00 \\
    \hline
  \end{tabular}
  \caption{\small \it Moments of Selected Assets' returns. For each return
    series, the 1st row contains the sample moments, while the 2nd
    constains those computed with MLE parameters. The time series are
    30-minute returns and run from 2013/10/10 to 2014/03/12.}
  \label{tab:assets_moments}
\end{table}

\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5, clip=true, trim=50 233 63
  136]{../pics/Volvo_B_30m_returns.pdf}
  \caption{\small \it Volvo B 30min returns' unconditional distribution fit to
    the Stochastic Volatility model. The time series runs from
    2013/10/10 to 2014/03/12. Left: $P(r < x)$ with $x < 0$; Right:
    $P(r > x)$ with $x > 0$. Both plots are on log-log scale.}
  \label{fig:Volvo_B_30m_returns}
\end{figure}

\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5, clip=true, trim=65 241 72
  143]{../pics/Nordea_30m_returns.pdf}
  \caption{\small \it Nordea Bank 30min returns' unconditional
    distribution fit to the Stochastic Volatility model. The time
    series runs from 2013/10/10 to 2014/03/12. Left: $P(r < x)$ with
    $x < 0$; Right: $P(r > x)$ with $x > 0$. Both plots are on log-log scale.}
  \label{fig:Nordea_30m_returns}
\end{figure}

\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5, clip=true, trim=21 241 28
  146]{../pics/Ericsson_30m_returns.pdf}
  \caption{\small \it Ericsson B 30min returns' unconditional
    distribution fit to the Stochastic Volatility model. The time
    series runs from 2013/10/10 to 2014/03/12. Left: $P(r < x)$ with
    $x < 0$; Right: $P(r > x)$ with $x > 0$. Both plots are on log-log scale.}
  \label{fig:Ericsson_30m_returns}
\end{figure}

>>>>

\section{GARCH models} \label{sec:Garch_model}
``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A GARCH(p, q) model is defined by the following
equation system:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu + \epsilon_t \\
    \sigma_t^2 &= \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
    \sum_{i=1}^p \beta_i \sigma_{t-i}^2
  \end{aligned}
\end{equation}
where $\epsilon_t$ is a random variable of a prescribed distribution,
the most common ones being Gaussian, Student's t, Asymmetric Student's
t, etc; $\sigma_t^2$ is the variance of $\epsilon_t$
\cite{Bollerslev86}. The autocorrelation function of $\epsilon_t^2$ is
given by (c.f. \cite{Bollerslev87})
\begin{eqnarray*}
\rho_n &=& \sum_{i=1} ^{\max(p, q)} (\alpha_i + \beta_i) \rho_{n-i}
\text{ for $n > p$}
\end{eqnarray*}
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. It is clear from this equation that the partial autocorrelation
function cuts off at $\max(p, q)$.

@@

Thus one can try a few frequencies and compare the distribution of
$r_{t+h, h}/\hat{\sigma}_{t+h, h}$ with the standard Gaussian. If the
two match, $\hat{\sigma}_{t+h, h}$ is a good estimate of $\sigma_{t+h,
  h}$, the volatility of $r_{t+h, h}$.

@@

As another measure of the quality of forecast, the fraction of
``good'' forecasts that differ no more than a given percentage from
the measured value is worth a few words. Table
\ref{tab:nordea_good_percentage} compares this fraction of good
forecasts from the 3 alternatives with respect to different
requirements of goodness.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ &
    ARIMA & GARCH & Sample Mean \\
    \hline
    1\% & 0.1920 & 0.2010 & 0.1843 \\
    \hline
    5\% & 0.7887 & 0.7436 & 0.7229 \\
    \hline
    10\% & 0.9665 & 0.9601 & 0.9369 \\
    \hline
  \end{tabular}
  \caption{\small \it Fraction of ``good'' forecasts as defined by
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln
      \hat{\sigma}_t|}$ being less than 1\%, 5\% and 10\%.}
  \label{tab:nordea_good_percentage}
\end{table}
Clearly, although ARIMA and GARCH rank differently with respect to
different requirements of goodness, they both out-perform the sample
mean forecast. So the efforts of building models and estimating them
have paid off.
\subsection{Volvo B Log-Volatility}\label{sec:volvo}
In this section we model the log-volatility of {\it Volvo B} 30-minute
returns during the period 2013/10/10 - 2014/04/04. This series
contains 1884 log-volatilities computed using 2-minute returns in each
30-minute interval. Among them we use the first 1507 for model
estimation and the last 377 for forecast and model verification.

@@

Following the aforementioned procedure, i.e. computing the initial
estimate by matching the moments and then refining the estimate by
MLE, we obtain the parameter values for a number of return series,
including the Volvo B 30-minute returns described in
\S\ref{sec:SLV_Symmetric}. Table \ref{tab:assets_params} and
\ref{tab:assets_moments} present the obtained parameter values and the
resulting moments, respectively. Figure \ref{fig:Volvo_B_30m_returns},
\ref{fig:Nordea_30m_returns}, and \ref{fig:Ericsson_30m_returns}
compare the empirical distribution functions with the analytic
distribution functions evaluated at these parameters.

Clearly these figures show a fairly good match of the empirical and
the analytic distribution functions. However, we also see that the
skewness of the returns of Volvo B and Ericsson B have rather
different values when computed from the sample and from the
model. This could be the consequence of the limited sample size or
deficiencies in the estimation procedure described above. We leave
these issues to future studies.

@@
